---
title: "AprilM_11 Final Project"
author: "April Meyer"
date: "2/21/2020"
output: 
  word_document: default
---

```{r setup, include=FALSE}
library(readr)
library(ggplot2)  # For plotting
library(class)
library(VIM)
library(tidyr)
library(stringr)
library(dplyr)
library(usmap)
```
#### Introduction  
After the 2014 killing of Michael Brown in Ferguson, Missouri, The Washington Post has started to collect a database of every fatal shooting in the US by a police officer in the line of duty. The protest movement, Black Lives Matter has increased focus on police accountability. This data includes information about the deceased including race, age, gender, if they were armed, mental health, fleeing and location. The post does not track deaths of people in police custody, non-shooting deaths, or by off-duty police. They aim to document conditions most like the Michael Brown in Ferguson. Centers for Disease control and FBI track fatal shootings by police, but according to Washington Post their data is incomplete. There has not been much dependable data around shootings by police before this movement and I would like to understand the current trends among fatal police shootings.  

#### The Problem Statement  
After the fatal shooting in 2014 waves of public protest broke out and lead to the movement of Black Lives Matter. I aim to look at the Washington Post data to see if there are trends that show signs of biases.  

#### Analysis  
My analysis included importing and cleaning the data. Once that was complete I sliced the data sets to compare variables.  
```{r import and clean}
#importing into a data frame
PoliceKillingsUS <- read.csv("PoliceKillingsUS.csv")
#Starting to clean and manipulate data
# Replace all empty strings in flee & race with NA
PoliceKillingsUS$flee[PoliceKillingsUS$flee == ""] <- NA
PoliceKillingsUS$race[PoliceKillingsUS$race == ""] <- NA
PoliceKillingsUS$armed[PoliceKillingsUS$armed == ""] <- NA
#remove all rows with any missing values
PoliceKillingsUS_clean <- na.omit(PoliceKillingsUS)
# Apply separate() to date
PoliceKillingsUS_clean <- separate(PoliceKillingsUS_clean, col = date, into = c("Day", "Month", "Year" ), sep = "/")
PoliceKillingsUS_clean$Month <- as.factor(PoliceKillingsUS_clean$Month)
PoliceKillingsUS_clean$Year <- as.factor(PoliceKillingsUS_clean$Year)
#cleaning up name column by converting to character and trimming
PoliceKillingsUS_clean$name <- as.character(PoliceKillingsUS_clean$name)
PoliceKillingsUS_clean$name <- str_trim(PoliceKillingsUS_clean$name) 
#cleaning up city & state column by converting to character and trimming
PoliceKillingsUS_clean$name <- as.character(PoliceKillingsUS_clean$city)
PoliceKillingsUS_clean$name <- str_trim(PoliceKillingsUS_clean$city) 
PoliceKillingsUS_clean$name <- as.character(PoliceKillingsUS_clean$state)
PoliceKillingsUS_clean$name <- str_trim(PoliceKillingsUS_clean$state) 
# Apply unite() to PoliceKillingsUS_clean
PoliceKillingsUS_cc <- unite(PoliceKillingsUS_clean, city_st, city, state, sep = ", ")
PoliceKillingsUS_cc$city_st <- as.factor(PoliceKillingsUS_cc$city_st)
# Check the structure
str(PoliceKillingsUS_cc)
```  

```{r summary}
# View a summary
summary(PoliceKillingsUS_cc)
```    

```{r slice and dice}
#creating different variables female and male
femaleSubset <- subset(PoliceKillingsUS_cc, subset = gender == "F")
maleSubset <- subset(PoliceKillingsUS_cc, subset = gender == "M")

#creating different variables bodyCamera
noCameraSubset <- subset(PoliceKillingsUS_cc, subset = body_camera == FALSE)
CameraSubset <- subset(PoliceKillingsUS_cc, subset = body_camera == TRUE)

#creating different variables mental illness
noSignsMentalSubset <- subset(PoliceKillingsUS_cc, subset = signs_of_mental_illness == FALSE)
SignsMentalSubset <- subset(PoliceKillingsUS_cc, subset = signs_of_mental_illness == TRUE)

#creating different variables mental illness
LosAngelesSubset <- subset(PoliceKillingsUS_cc, subset = city_st == "Los Angeles, CA")
PhoenixSubset <- subset(PoliceKillingsUS_cc, subset = city_st == "Phoenix, AZ")
``` 

#### Implications   
I have done some plots to show how variables compare and totals for some. I have also attempted to fit a glm model to see if a person is armed. Age and signs of mental illness would be worth looking into further. Since they have the greatest effect because the Pr(>|z|) is less than 0.05 for those four variables. 

```{r plots, echo=FALSE}
#table of armed counts
# tableArmed <- table(PoliceKillingsUS_cc$armed)
# as.data.frame(tableArmed)
na.omit(PoliceKillingsUS_cc) %>% 
  group_by(PoliceKillingsUS_cc$armed) %>%
  summarize(Count=n()) %>%
  mutate(Percent = round((Count/sum(Count)*100))) %>%
  arrange(desc(Count))
#table of armed counts
na.omit(PoliceKillingsUS_cc) %>% 
  group_by(PoliceKillingsUS_cc$city_st) %>%
  summarize(Count=n()) %>%
  mutate(Percent = round((Count/sum(Count)*100))) %>%
  arrange(desc(Count))
#bar chart Armed Individuals in Phoenix
ggplot(PhoenixSubset) + geom_bar(aes(x = armed)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + labs(title="Armed Individuals in Phoenix")

#bar chart Mental Illness by Gender
ggplot(PoliceKillingsUS_cc) + geom_bar(aes(x = signs_of_mental_illness)) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + facet_wrap(~ gender) + labs(title="Mental Illness by Gender")

#usMap of deaths
st_count <- summary(PoliceKillingsUS_clean$state)
st_count <- stack(st_count)
st_count <- st_count[c(2,1)]
st_count$fips <- fips(st_count$ind)
names(st_count)[2] <- "Total"
plot_usmap(data = st_count, values = "Total", color = "blue", labels = FALSE) + 
  scale_fill_continuous(low = "white", high = "blue", name = "Total", label = scales::comma) +
  theme(legend.position = "right") + 
  theme(panel.background = element_rect(colour = "black")) +
  labs(title = "Deaths by State in the US")

#line chart by year
yr_count <- PoliceKillingsUS_cc %>%
  group_by(Month, Year) %>%
  summarise(count=n())
ggplot(data = yr_count, aes(y=count, x = Month, group = Year, colour = Year)) + geom_line()
```

```{r predicting if armed, echo=FALSE}
#Starting to clean and manipulate data
PoliceKillingsUS_armed <- PoliceKillingsUS[c("armed", "age", "gender", "race", "signs_of_mental_illness", "flee")]
# Replace all empty strings in flee & race with NA
PoliceKillingsUS_armed$flee[PoliceKillingsUS$flee == ""] <- NA
PoliceKillingsUS_armed$race[PoliceKillingsUS$race == ""] <- NA
PoliceKillingsUS_armed$armed[PoliceKillingsUS$armed == ""] <- NA
#remove all rows with any missing values
PoliceKillingsUS_armed <- na.omit(PoliceKillingsUS_armed)
PoliceKillingsUS_armed$armed <- as.character(PoliceKillingsUS_armed$armed)
PoliceKillingsUS_armed$armed <- str_trim(PoliceKillingsUS_armed$armed) 
# Converting armed to 0 is armed and 1 is not armed
PoliceKillingsUS_armed$armed[PoliceKillingsUS_armed$armed == "undetermined"] <- "unarmed"
PoliceKillingsUS_armed$armed[PoliceKillingsUS_armed$armed != "unarmed"] <- "1"
PoliceKillingsUS_armed$armed[PoliceKillingsUS_armed$armed == "unarmed"] <- "0"
PoliceKillingsUS_armed$armed <- as.integer(PoliceKillingsUS_armed$armed)
# Check the structure
str(PoliceKillingsUS_armed)

model_all <- glm(formula = armed ~ ., family = binomial(), data = PoliceKillingsUS_armed)
summary(model_all)
```  
#### Limitations  
There were many limitations to this data set. First once I started digging into the data more there were many undefined, missing, and NA values. Also, there were inaccuracies in data due to duplicates. The source of the data caused problems; for example, using only existing newspaper sources which could not include small towns or shootings that did not make the newspapers. Lastly it was harder to do any predictive modelling due to all the categorical data.  
#### Concluding Remarks  
I enjoyed working on this project; however, I do wish I would have picked a better data set. I did not realize how many limitations I would encounter with this set. Furthermore, I want to learn more on how to prepare and clean data. That would have been helpful with this data set. 